Towards Transparency and Open Science

### 2023-01-16 manuscript.Rmd
single space toc
first draft of summary of reproducibility

\tableofcontents

@@

The need for transparency is closely tied to the use of induction in the empirical test of a theory.
There has been a long and vigorous debate about what it means to test a theory empirically [@popperLogicScientificDiscovery2002; Ch. 5 Experience as a Method].
I do not attempt to rehash the debate about what constitutes an empirical test but aim to lay open the role of transparency in two frameworks that lend themselves to investigate induction.
The first framework motivates transparency when the aim of an empirical test is to evaluate the verisimilitude ("closeness to truth") of a theory.
The second framework motivates transparency under a science that wants to select a theory according to its expected predictive performance.

@@

The above sections aimed to motivate the observation that the apparent fit of a theory to data is often overly optimistic, if it has inductive elements.
This observation is only of limited use if we do not now the extend of this optimism.
However, both setups show that the extent of the optimism ($\mathcal{C}$) is closely related to the inductive process and give us two starting points for making this bias transparent.
The first requires a formal analysis of the inductive process to compute the complexity (e.g., using information criteria), and the second merely requires that the process is repeatable (e.g., using cross-validation).
Both approaches require researchers to make the process of induction transparent instead of merely publishing the results.

@@

It is without question that researchers sometimes engage in inductive behavior that is neither formally analyzable nor repeatable.
This fact implies that for theses situations the optimism bias can not be fully quantified and that full transparency can not be archived.
To enable proper judgment of the whole theory, the imperative is simple: induce only what is necessary and what you induce should, if at all possible, be done formally.
Otherwise, the supposedly objective test of the theory using hard data must still be judged subjectively.

@@

In the following, I propose preregistration, as means to move induction into the formal domain, and computational reproducibility to make formal induction transparent.

@@

The following article is reprinted from the following source under [Creative Commons Attribution (CC BY) 4.0 International License](https://creativecommons.org/licenses/by/4.0).

@@

The following article is prepared to be submitted to *Philosophical Psychology*.
The preprint is reprinted here under [Creative Commons Public Domain Dedication (CC0 1.0)](https://creativecommons.org/publicdomain/zero/1.0/).

>>>>

\begin{spacing}{1}
\tableofcontents
\end{spacing}

@@

The need for transparency is closely tied to the use of induction in the empirical test of a theory.
There has been a long and vigorous debate about what it means to test a theory empirically [@popperLogicScientificDiscovery2002, Ch. 5 Experience as a Method].
I do not attempt to rehash the debate about what constitutes an empirical test but aim to lay open the role of transparency in two frameworks that lend themselves to investigate induction.
The first framework motivates transparency when the aim of an empirical test is to evaluate the verisimilitude ("closeness to truth") of a theory.
The second framework motivates transparency under a science that wants to select a theory according to its expected predictive performance.

@@

The above sections aimed to motivate the observation that the apparent fit of a theory to data is often overly optimistic, if it has inductive elements.
This observation is only of limited use if we do not know the extend of this optimism.
However, both setups show that the extent of the optimism  is closely related to the complexity of the inductive process ($\mathcal{C}$), and give us two starting points for making this bias transparent.
The first requires a formal analysis of the inductive process to compute the complexity (e.g., using information criteria), and the second merely requires that the process is repeatable (e.g., using cross-validation).
Both approaches require researchers to make the process of induction transparent instead of merely publishing the results.

@@

What we can do is to split $\mathcal{C}$ in complexity that can be formally described and a part that can only be evaluated subjectively.
Formal, in the strict sense means that the complexity can be mathematically derived and in a looser sense that the process can be repeated at will.
If a researcher employs a linear model, the complexity is calculable and even if it were not, we could fit the linear model on a large set of other data to asses its inductive behavior.
However, if the researcher reconsiders the model based on the results, and fits a model with, e.g., a predictor added or removed, the complexity can not be formally judged because the induction is partly done by the researcher themselves.
We can not ask the researcher to repeat the process on all possible data sets nor is the behavior mathematically well defined.
What we know, however, is that the resulting linear model is more complex than a linear regression formally implies.

$$
\mathcal{C}(\text{Model}) = \mathcal{C}(\text{formal}) +\mathcal{C}(\text{informal})
$$

That is not to say that we have no basis to judge the informal induction.
We can find the inductive decision reasonable and judge it unlikely that just about any variable would be added if the data suggests it.
Or the opposite, we might find it not well justified on theoretical grounds and deem it a purely data driven decision, which implies higher complexity.
What one can say, however, is that this judgment is debatable and, therefore, subjective.

It is without question that researchers sometimes engage in inductive behavior that is neither formally analyzable nor repeatable.
This fact implies that for theses situations the complexity and hence the optimism bias can not be fully quantified.
The second best thing to transparency about the complexity is to communicate the uncertainty about the complexity clearly.
However, judging the informal complexity $\mathcal{C}(\text{informal})$ must remain an subjective exercise.

To enable proper judgment of the whole theory, the imperative is simple: induce only what is necessary and what you induce should, if at all possible, be done formally.
Otherwise, the supposedly objective test of the theory using hard data must be judged more subjectively than necessary.

To summarise, I separate complexity $\mathcal{C}$ according to the transparency that can theoretically be achieved.
Formal inductive processes allow full transparency in the sense that $\mathcal{C}$ can be objectively quantified.
Informal inductive processes allow only a limited transparency and must be subjectively judged.

@@

In the following, I propose preregistration, as means to move induction into the formal domain, and computational reproducibility to make formal induction transparent.
I argue that both practices are ultimately aimed at providing transparency to judge empirical support for theories, though they may have secondary benefits for science.

Assume for a moment that a researcher only engaged in formal induction while testing a theory empirically.
If other researchers want to judge this empirical test properly, the inductive process must be communicated to them.
In other words, having formal induction is not enough, the induction still must be made transparent.
Such transparency can be achieved by computational reproducibility.
Computational reproducibility is usually defined as the ability to recreate the same results from the same data set.
However, through the lens of transparency about the inductive process, we may further refine our view on computational reproducibility.

Consider a software that delivers the same results upon input of the same data, i.e., its reproducible.
However, place two restrictions on the software.
First, while able to compute results based on data, the software can not be understood by human researcher, e.g., because it is a binary program.
Second, when the data set changes, the software does not work anymore.
The first condition rules out that the software can be analyzed formally.
The second condition prevents us from changing the data in any way and observe the results.
We, therefore, have no means to assess the complexity of the inductive process and can not judge the empirical support.
It follows that to provide transparency about the complexity of an inductive process the definition of computational reproducibility must be broadened, i.e., we have to rule out one of the above conditions.

If we relax the second condition and have a "black box" that is nerveless repeatable on other data, we still have transparency about the inductive process, in the sense, that we can estimate the complexity of the inductive process.
Relaxing the first condition most likely leads to relaxing the second condition as well.
If the inductive process can be understood by humans, complexity can either be calculated analytically or the process can be reimplemented to work on other similar data.
The most minimal extension of computational reproducibility is, therefore, to require additionally that the process can be repeated on other similar data.

To provide transparency, computational reproducibility must satisfy two requirements.
First, computational reproducibility must show that the results have been the product of the proposed inductive process, i.e., the same data leads to the same results.
Second, computational reproducibility must enable that the inductive process can be repeated on similar data.

These two requirements are not easy to meet in practice.
In @peikertReproducibleDataAnalysis2021, I proposed a workflow that unifies both requirements under the objective of automating the full process from data to manuscript.
Removing the human from the loop enables to formally verify that the first requirement is met, and drastically reduces the burden of the second requirement.
While this workflow provides transparency and facilitates that the recipients of the work can judge the inductive process easily, it asks a lot from the researchers who create the work.
To address this problem, and make the application of the workflow easier I developed the R package `repro` [@R-repro].
I further refined and simplified the workflow to make it more accessible in @peikertReproducibleResearchTutorial2021 and contributed to @vanlissaWORCSWorkflowOpen2021's manuscript and the there presented software `worcs`.

To address the problem of making transparent how the data led to the results, these works divide it into four sub problems.
First, it must be unambiguous what results are generated by which inductive process.
This can be ensured by employing dynamic document creation.
Second, the version of the software that is used to generate the results must be known.
This can be ensured by documenting the version using software management.
Third, the version of the author written code (and possibly text) must be tracked using version control systems.
Fourth, how the data relates to software and computing infrastructure must be managed by e.g. workflow automation.

Addressing these four problems through automation lowers the barrier to understand, and hence make transparent, the inductive process.
However, it also addresses the here proposed additional requirement for computational reproducibility, that the process can be repeated on similar data, as it effectively allows to scale the process across many data sets.

## Transparency about human researchers: Preregistration

### Theorethical

<!--you must pay the reaper-->
Accounting for how much induction was used to deductively test a statement is necessary for correct appraisal of evidential support of a hypothesis.
Researcher (and philosophers of science) differ considerably in what appraisal entails, e.g., judging the long-term performance (if they are frequentists), degrees of believe (if they are Bayesians), or probativeness [if they are severe testers, @mayoStatisticalInferenceSevere2018, p. 14] of a hypothesis.
But whatever measure they subscribe to, induction can shield shield any hypothesis from refutation by data altogether.

Assume that all auxiliary assumption, regarding proper quantification of behavior and situation in valid measurements, causality, individual differences, nuisance variables, distributional, etc, hold.

While not using statistical methods would lead to the rejection of almost all psychological theories, pushed to the extreme statistical methods can shield theories from refutation altogether.
Both situations are uninformative and, therefore, rather unhelpful for the generation of knowledge.
The later situation, however, becomes less than helpful when one is under the impression that theories are well tested, when in fact there was little chance that the data could refute the theory if wrong.

@@

The following article is reprinted from the following source under the conditions of the [Creative Commons Attribution (CC BY) 4.0 International License](https://creativecommons.org/licenses/by/4.0).

@@

The following article is prepared to be submitted to *Philosophical Psychology*.
The following preprint is reprinted from the following source under the conditions of the [Creative Commons Public Domain Dedication (CC0 1.0)](https://creativecommons.org/publicdomain/zero/1.0/).
